---
title: "Analyses"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
#library(xlsx)
library(tidyr)
library(dplyr)
library(ggplot2)
library(haven)
library(car)
library(ez)
library(lme4)
library(lmerTest)

```

## TODO discussion 2021-09-20

- why negative: because inverted u and a in 2 places, see chunk calc-space
- why different dfs: I wasn't doing the same analysis, now I am
- why different Fs: because the vss data was different (due to formula switch)
- explain about 2 rows that are excluded, & possible mismatch
- do we have pitch.txt & formant.txt outputted by script?
  - no, but we have output by mom! Alex to integrate
- we can also leave that be, since by now we are aligned in terms of means for point vowels?
  - no, because duration & pitch are incorrectly attributed
- not sure what to do about the fact that our vowel triangles don't look the same -- how were those generated? perhaps they are not over the same tokens?
  - mystery solved! i and u based on all moms in both figures
  - what shall we do, single figure for all moms?
- "REAL" decisions:
  - removing outliers? -- yes
  - removing items from f1/2 when no f0 -- keeping them because they are not outliers
  - doing ANOVA first, then mixed model- agreed
  - (I don't think software matters!) - ANOVA in spss, lmer in r

### interim done:
- switched input file, results are stable
- add confidence intervals to vowel space, incorporate results to paper


## Replication

In this section, I'm just doing again the analyses that are already in the [manuscript](https://docs.google.com/document/d/1iQkz4HziQi6ECLrSMRTXtzBQyt29a5Ep/edit). Those were done with SPSS, so this is more replication than reproduction.

```{r readin}

#read.xlsx("Full Puppy Data Set_summarized.xls",1)->dat
#read.csv("input/Full Puppy Data Set_summarized/Sheet1-Table 1.csv")->dat
#read.csv("input/Puppy Data Raw/Sheet1-Table 1.csv")->dat
#read.csv("input/MODIFIED_Full Puppy Data Set_Sep 21/Sheet1-Table 1.csv")->dat
read.csv("input/Current Data Set_Sep 30/Modified-Table 1.csv")->dat
#remove empty rows
dat=dat[!is.na(dat$Mom) &dat$Mom !="" ,]
dim(dat) #614 obs

#excluded data
#dat[!(dat$Mom %in% c( " Mom2ADSbox5beadv.wav "," Mom2ADSbox5bootv.wav " )),]->dat


dat$Mom=tolower(gsub(" ","",dat$Mom)) #clean up
#split out info from Mom column
dat$motID=gsub("_.*","",dat$Mom)
dat$motID=gsub(".ds.*","",dat$motID)
#table(dat$motID)

#split out DS info
dat$ds=NA
dat$ds[grep("pd",dat$Mom)]<-"pds"
dat$ds[grep("ads",dat$Mom)]<-"ads"
dat$ds[grep("_ad_",dat$Mom)]<-"ads"
dat$ds[grep("id",dat$Mom)]<-"ids"
#table(dat$ds)
# sum(table(dat$ds))

#split out vowel info
dat$v=NA
dat$v[grep("bead",dat$Mom)]<-"i"
dat$v[grep("boot",dat$Mom)]<-"u"
dat$v[grep("ball",dat$Mom)]<-"a"
dat$v[grep("box",dat$Mom)]<-"a"
dat$stim_group<-"box"
dat$stim_group[dat$motID %in% c("mom7","mom8","mom9","mom10")]<-"ball"

dat$target=NA
dat$target[grep("bead",dat$Mom)]<-"bead"
dat$target[grep("boot",dat$Mom)]<-"boot"
dat$target[grep("ball",dat$Mom)]<-"ball"
dat$target[grep("box",dat$Mom)]<-"box"

```

Checking that we are getting the same means against the means in the xls file -- we are.

```{r chk-means-xls, eval=F}
excel_avg=dat[!is.na(dat$F1.mean),c("Mom","F1.mean","F2.mean","motID","ds","v")]
excel_avg$uID=paste(excel_avg$motID,excel_avg$ds,excel_avg$v)
means$uID=paste(means$motID,means$ds,means$v)
merge(excel_avg,means,by="uID")->test
test[round(test$F1.mean)!=round(test$f1),c("Mom","F1.mean","f1")]
test[round(test$F2.mean)!=round(test$f2),c("Mom","F2.mean","f2")]

```

Checking that we are getting the same means against the means in the sav file -- we are.


```{r chk-means-sav,eval=F}
read_sav("input/Working Data File Aug 25 2021.sav")->spss
write.csv(spss,"output/spss.csv")
spss[!is.na(spss$ADBallF1),]->spss #remove empty lines
spss$motID<-tolower(gsub(" ","",spss$V1)) #clean motID
spss[,-1]->spss

pivot_longer(spss,-motID,values_to="value",names_to = "type")->long
long$ds=long$v=long$f=NA
long$ds[grep("AD",long$type)]<-"ads"
long$ds[grep("ID",long$type)]<-"ids"
long$ds[grep("PD",long$type)]<-"pds"
long$v[grep("Ball",long$type)]<-"a"
long$v[grep("Bead",long$type)]<-"i"
long$v[grep("Boot",long$type)]<-"u"
long$f[grep("F1",long$type)]<-"f1"
long$f[grep("F2",long$type)]<-"f2"
long[!is.na(long$f),c("motID","ds","v","f","value")]->long

pivot_longer(means,f1:f2,names_to="f",values_to = "value")->means_long

# sort and compare
means_long$unique=paste(means_long$motID,means_long$ds,means_long$v,means_long$f)
long$unique=paste(long$motID,long$ds,long$v,long$f)
sum(round(means_long[order(means_long$unique),"value"])==round(long[order(long$unique),"value"]))

```


Note there are no outliers:


```{r}
ggplot(dat, aes(x=f1, y=f2, color=ds)) +
  geom_text(label=dat$v)
```


```{r outliers and means}

table(dat$target)
#old_dat[old_dat$Mom %in% outliers,]
dim(dat) #612 obs

#table(dat$v)
# sum(table(dat$v))

dat %>%
    group_by(motID,ds,v,stim_group,target) %>%
    dplyr::summarize(f1 = mean(f1, na.rm=TRUE),
                     f2 = mean(f2, na.rm=TRUE)) -> means


means %>%
    group_by(ds,v,stim_group) %>%
    dplyr::summarize(f1 = mean(f1, na.rm=TRUE),
                     f2 = mean(f2, na.rm=TRUE)) -> overall_means

dat %>%
    group_by(motID,ds,target) %>%
    dplyr::summarize(f1 = mean(f1, na.rm=TRUE),
                     f2 = mean(f2, na.rm=TRUE)) -> means_target
means_target %>%
    group_by(ds,target) %>%
    dplyr::summarize(f1 = mean(f1, na.rm=TRUE),
                     f2 = mean(f2, na.rm=TRUE)) -> overall_means_target

means %>%
    group_by(ds,target) %>%
    dplyr::summarize(f1 = sd(f1, na.rm=TRUE),
                     f2 = sd(f2, na.rm=TRUE)) -> sds

means %>%
    group_by(ds,v) %>%
    dplyr::summarize(f1 = mean(f1, na.rm=TRUE),
                     f2 = mean(f2, na.rm=TRUE)) -> overall_means_nogroup



```

## Fig vowel space


```{r fig-all}
p <- ggplot(subset(overall_means_nogroup), aes(x = f1, y = f2, col = factor(ds)))+
  geom_polygon(data = subset(overall_means_nogroup) ,
  aes(
     x=f1 ,
    y=f2 ,
    group=ds
  ), fill=NA) + #alpha = 0.3 (if we want fill)
  geom_point() +
  annotate("text", x= 350,y=2800,label="i") +
  annotate("text", x= 850,y=1350,label="a") +
  annotate("text", x= 350,y=1350,label="u")
 
pdf("fig1.pdf")
p
dev.off()

```

```{r fig-all2}

overall_means_target$target=factor(overall_means_target$target, levels=c("bead","box","ball","boot"), ordered=T)
overall_means_target[order(overall_means_target$target),]->overall_means_target

sds$target=factor(sds$target, levels=c("bead","box","ball","boot"), ordered=T)
sds[order(sds$target),]->sds

p <- ggplot(subset(overall_means_target), aes(x = f1, y = f2, col = factor(ds)))+
  geom_polygon(data = subset(overall_means_target) ,
  aes(
     x=f1 ,
    y=f2 ,
    group=ds
  ), fill=NA) + #alpha = 0.3 (if we want fill)
  geom_point() 
p 



```

```{r}
p +  geom_errorbar(aes(ymin = overall_means_target$f2 - sds$f2,ymax = overall_means_target$f2 + sds$f2)) + 
    geom_errorbarh(aes(xmin = overall_means_target$f1 - sds$f1,xmax = overall_means_target$f1 + sds$f1)) +ggtitle("1 SDs")
```


```{r}
p +  geom_errorbar(aes(ymin = overall_means_target$f2 - 2*sds$f2,ymax = overall_means_target$f2 + 2*sds$f2)) + 
    geom_errorbarh(aes(xmin = overall_means_target$f1 - 2*sds$f1,xmax = overall_means_target$f1 + 2*sds$f1)) +ggtitle("2 SDs")
```


## Vowel space calculations

Next we calculate vowel space for each mom and register.
```{r calc-space}

#formula from footnote
# (F1/i/*(F2/u/ - F2/ɑ/) + F1/u/*(F2/i/ - F2/ɑ/) + F1/ɑ/*(F2/i/ - F2/u/))/2
#  with(means, vowelMeansPolygonArea(f1, f2, vowel=v, poly.order=c("i","a","u"), group="motID"))
# 
# areapl(overall_means$f1,overall_means$f2)

#formula from footnote
# (F1/i/*(F2/u/ - F2/ɑ/) + F1/u/*(F2/i/ - F2/ɑ/) + F1/ɑ/*(F2/i/ - F2/u/))/2
# from syntax file
# COMPUTE IDVowelSpace1 = (IDBeadF1 * (IDBoxF2-IDBootF2) + IDBoxF1 * (IDBootF2
#  -IDBeadF2)+IDBootF1*(IDBeadF2-IDBoxF2))/2 .
# EXECUTE .
# COMPUTE PDVowelSpace1 = (PDBeadF1 * (PDBoxF2-PDBootF2) + PDBoxF1 * (PDBootF2
#  -PDBeadF2)+PDBootF1*(PDBeadF2-PDBoxF2))/2 .
# EXECUTE .
# COMPUTE ADVowelSpace1 = (ADBeadF1 * (ADBoxF2-ADBootF2) + ADBoxF1 * (ADBootF2
#  -ADBeadF2)+ADBootF1*(ADBeadF2-ADBoxF2))/2 .
# EXECUTE .
# COMPUTE IDVowelSpace2 = (IDBeadF1 * (IDBallF2-IDBootF2) + IDBallF1 * (IDBootF2
#  -IDBeadF2)+IDBootF1*(IDBeadF2-IDBallF2))/2 .
# EXECUTE .
# COMPUTE PDVowelSpace2 = (PDBeadF1 * (PDBallF2-PDBootF2) + PDBallF1 * (PDBootF2
#  -PDBeadF2)+PDBootF1*(PDBeadF2-PDBallF2))/2 .
# EXECUTE .
# COMPUTE ADVowelSpace2 = (ADBeadF1 * (ADBallF2-ADBootF2) + ADBallF1 * (ADBootF2
#  -ADBeadF2)+ADBootF1*(ADBeadF2-ADBallF2))/2 .
# EXECUTE .
vs=NULL
for(thismom in levels(factor(dat$motID))) for(thisds in levels(factor(dat$ds))){
  mydat=subset(means,means$motID==thismom & means$ds==thisds)
  vs=rbind(vs,
           cbind(
# (F1/i/*(F2/u/ - F2/ɑ/) + F1/u/*(F2/i/ - F2/ɑ/) + F1/ɑ/*(F2/i/ - F2/u/))/2
      (mydat$f1[mydat$v=="i" & mydat$motID==thismom]*
         (mydat$f2[mydat$v=="u" & mydat$motID==thismom] - mydat$f2[mydat$v=="a" & mydat$motID==thismom]) +
        mydat$f1[mydat$v=="u" & mydat$motID==thismom]*
         (mydat$f2[mydat$v=="i" & mydat$motID==thismom] - mydat$f2[mydat$v=="a" & mydat$motID==thismom]) +
        mydat$f1[mydat$v=="a" & mydat$motID==thismom]*
         (mydat$f2[mydat$v=="i" & mydat$motID==thismom] - mydat$f2[mydat$v=="u" & mydat$motID==thismom])
      )/2,
          thismom,thisds )
        )
}

colnames(vs)<-c("vss","mom","ds")
write.csv(vs,"output/vs.csv",row.names=F)

read.csv("output/vs.csv")->vs


pivot_wider(vs,names_from="ds",values_from="vss")->vs_wide
write.csv(vs_wide,"output/vs_wide.csv",row.names=F)

vs %>%
    group_by(ds) %>%
    dplyr::summarize(vss = mean(vss, na.rm=TRUE)) 
```

## ANOVA

A within-subject analysis of variance (ANOVA) on average vowel space across
listener conditions (3=IDS, PDS, ADS) revealed a significant main effect (F(2,18) = 6.04,
p &lt; .02), with paired contrasts indicating that the average vowel spaces for IDS and
PDS were not significantly different, t(9) = .98, p &gt; .05, but were both greater than the
vowel space of ADS (IDS v. ADS, t(9) = 3.79, p &lt; .005) and (PDS v. ADS, t(9) = 2.24, p
= .05; see Figure 1a and b).

ANOVA: not the same F or dfs

explanations from Robin: "To conduct a repeated-measures ANOVA in SPSS, we do not specify the repeated-measures factor and the dependent variable in the SPSS data file. Instead, the SPSS data file contains several quantitative variables. The number of quantitative variables is equal to the number of levels of the within-subjects factor. The scores on any one of these quantitative variables are the scores on the dependent variable for a single level of the within-subjects factor. Although we do not define the within-subjects factor in the SPSS data file, we specify it in the dialog box for the General Linear Model Repeated-Measures procedure. To define the factor, we give a name to the within-subjects factor, specify the number of levels of this factor, and indicate the quantitative variables in the data set associated with the levels of the within-subjects factor.  " and "the error term in the SPSS repeated measures model is (n-1) for sample (10-1=9) x (k-1) for the number of measures per sample (3-1=2);  so 9 x 2 = 18 (for the denominator)."

```{r}

#https://mikebyrnehfhci.wordpress.com/2015/08/03/translating-spss-to-r-mixed-repeated-measures-anova/
#https://rpubs.com/tomhardwicke/anova-check
mylm = lm(vss~ds + (1/mom),data=vs)
Anova(mylm,type=3)


ez_outcome=ezANOVA(vs, dv=vss, wid=mom, within=ds, detailed=T,
  return_aov = TRUE)
summary(ez_outcome)
model.tables(ez_outcome$aov, "means")


```



Paired contrasts -- same pattern of results but different t values.

```{r ts}
library(effsize)
t.test(vs$vss[vs$ds=="pds"],vs$vss[vs$ds=="ids"],paired=T)
cohen.d(vs$vss[vs$ds=="pds"|vs$ds=="ids"],vs$ds[vs$ds=="pds"|vs$ds=="ids"])
t.test(vs$vss[vs$ds=="ids"],vs$vss[vs$ds=="ads"],paired=T)
cohen.d(vs$vss[vs$ds=="ads"|vs$ds=="ids"],vs$ds[vs$ds=="ads"|vs$ds=="ids"])
t.test(vs$vss[vs$ds=="pds"],vs$vss[vs$ds=="ads"],paired=T)
cohen.d(vs$vss[vs$ds=="pds"|vs$ds=="ads"],vs$ds[vs$ds=="pds"|vs$ds=="ads"])

```

## dur, f0, f1, f2

Duration and pitch can be done at the item level

```{r}
mymod=lmer(Vowel.Duration~ds*target + (1 + ds|motID),data=dat)
summary(mymod)

dat$Pitch=as.numeric(as.character(dat$Pitch))
mymod=lmer(Pitch~ds*target + (1+ds|motID),data=dat)
summary(mymod)

```

```{r}
mymod=lmer(f1~ds*target + (1 + ds|motID),data=dat)
summary(mymod)

mymod=lmer(f2~ds*target + (1|motID),data=dat)
summary(mymod)
```


## Variability analyses

Get standard deviation of f1 and f2, separating by mom and register.

```{r}
dat %>%
    group_by(motID,ds,target) %>%
    dplyr::summarize(f1.sd = sd(f1, na.rm=TRUE),
                     f2.sd = sd(f2, na.rm=TRUE)) -> sds

```



```{r}
summary(lmer(f1.sd~ds*target + (1 |motID),data=sds))

```


```{r}
summary(lmer(f2.sd~ds*target + (1|motID),data=sds))

```



