---
title: 'Supplementary materials: Positive Valence Contributes to Hyperarticulation in Maternal Speech to Infants and Puppies'
output:
  pdf_document:
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F,warning=F,message = FALSE)
#library(xlsx)
library(tidyr)
library(dplyr)
library(ggplot2)
library(haven)
library(car)
library(ez)
library(lme4)
library(lmerTest)

```

## Initial data cleaning

```{r readin}

read.csv("input/puppyDS.csv")->dat
#remove empty rows
dat=dat[!is.na(dat$Mom) &dat$Mom !="" ,]
#dim(dat) #614 obs


dat$Mom=tolower(gsub(" ","",dat$Mom)) #clean up
#split out info from Mom column
dat$motID=gsub("_.*","",dat$Mom)
dat$motID=gsub(".ds.*","",dat$motID)
#table(dat$motID)

#split out DS info
dat$ds=NA
dat$ds[grep("pd",dat$Mom)]<-"pds"
dat$ds[grep("ads",dat$Mom)]<-"ads"
dat$ds[grep("_ad_",dat$Mom)]<-"ads"
dat$ds[grep("id",dat$Mom)]<-"ids"
#table(dat$ds)
# sum(table(dat$ds))

#split out vowel info
dat$v=NA
dat$v[grep("bead",dat$Mom)]<-"i"
dat$v[grep("boot",dat$Mom)]<-"u"
dat$v[grep("ball",dat$Mom)]<-"a"
dat$v[grep("box",dat$Mom)]<-"a"
dat$stim_group<-"box"
dat$stim_group[dat$motID %in% c("mom7","mom8","mom9","mom10")]<-"ball"

dat$target=NA
dat$target[grep("bead",dat$Mom)]<-"bead"
dat$target[grep("boot",dat$Mom)]<-"boot"
dat$target[grep("ball",dat$Mom)]<-"ball"
dat$target[grep("box",dat$Mom)]<-"box"

```

Note there are no outliers:


```{r}
ggplot(dat, aes(x=f1, y=f2, color=ds)) +
  geom_text(label=dat$v)
```


```{r outliers and means}

table(dat$target)

# dim(dat) #614 obs

#table(dat$v)
# sum(table(dat$v))

dat %>%
    group_by(motID,ds,v,stim_group,target) %>%
    dplyr::summarize(f1 = mean(f1, na.rm=TRUE),
                     f2 = mean(f2, na.rm=TRUE)) -> means


means %>%
    group_by(ds,v,stim_group) %>%
    dplyr::summarize(f1 = mean(f1, na.rm=TRUE),
                     f2 = mean(f2, na.rm=TRUE)) -> overall_means

dat %>%
    group_by(motID,ds,target) %>%
    dplyr::summarize(f1 = mean(f1, na.rm=TRUE),
                     f2 = mean(f2, na.rm=TRUE)) -> means_target
means_target %>%
    group_by(ds,target) %>%
    dplyr::summarize(f1 = mean(f1, na.rm=TRUE),
                     f2 = mean(f2, na.rm=TRUE)) -> overall_means_target

means %>%
    group_by(ds,target) %>%
    dplyr::summarize(f1 = sd(f1, na.rm=TRUE),
                     f2 = sd(f2, na.rm=TRUE)) -> sds

means %>%
    group_by(ds,v) %>%
    dplyr::summarize(f1 = mean(f1, na.rm=TRUE),
                     f2 = mean(f2, na.rm=TRUE)) -> overall_means_nogroup


means %>%
    group_by(ds,v) %>%
    dplyr::summarize(f1 = sd(f1, na.rm=TRUE),
                     f2 = sd(f2, na.rm=TRUE)) -> overall_sds_nogroup

```


Next we calculate vowel space for each mom and register.

```{r calc-space}

#formula from footnote
# (F1/i/*(F2/u/ - F2/ɑ/) + F1/u/*(F2/i/ - F2/ɑ/) + F1/ɑ/*(F2/i/ - F2/u/))/2
#  with(means, vowelMeansPolygonArea(f1, f2, vowel=v, poly.order=c("i","a","u"), group="motID"))
# 
# areapl(overall_means$f1,overall_means$f2)

#formula from footnote
# (F1/i/*(F2/u/ - F2/ɑ/) + F1/u/*(F2/i/ - F2/ɑ/) + F1/ɑ/*(F2/i/ - F2/u/))/2
# from syntax file
# COMPUTE IDVowelSpace1 = (IDBeadF1 * (IDBoxF2-IDBootF2) + IDBoxF1 * (IDBootF2
#  -IDBeadF2)+IDBootF1*(IDBeadF2-IDBoxF2))/2 .
# EXECUTE .
# COMPUTE PDVowelSpace1 = (PDBeadF1 * (PDBoxF2-PDBootF2) + PDBoxF1 * (PDBootF2
#  -PDBeadF2)+PDBootF1*(PDBeadF2-PDBoxF2))/2 .
# EXECUTE .
# COMPUTE ADVowelSpace1 = (ADBeadF1 * (ADBoxF2-ADBootF2) + ADBoxF1 * (ADBootF2
#  -ADBeadF2)+ADBootF1*(ADBeadF2-ADBoxF2))/2 .
# EXECUTE .
# COMPUTE IDVowelSpace2 = (IDBeadF1 * (IDBallF2-IDBootF2) + IDBallF1 * (IDBootF2
#  -IDBeadF2)+IDBootF1*(IDBeadF2-IDBallF2))/2 .
# EXECUTE .
# COMPUTE PDVowelSpace2 = (PDBeadF1 * (PDBallF2-PDBootF2) + PDBallF1 * (PDBootF2
#  -PDBeadF2)+PDBootF1*(PDBeadF2-PDBallF2))/2 .
# EXECUTE .
# COMPUTE ADVowelSpace2 = (ADBeadF1 * (ADBallF2-ADBootF2) + ADBallF1 * (ADBootF2
#  -ADBeadF2)+ADBootF1*(ADBeadF2-ADBallF2))/2 .
# EXECUTE .
vs=NULL
for(thismom in levels(factor(dat$motID))) for(thisds in levels(factor(dat$ds))){
  mydat=subset(means,means$motID==thismom & means$ds==thisds)
  vs=rbind(vs,
           cbind(
# (F1/i/*(F2/u/ - F2/ɑ/) + F1/u/*(F2/i/ - F2/ɑ/) + F1/ɑ/*(F2/i/ - F2/u/))/2
      (mydat$f1[mydat$v=="i" & mydat$motID==thismom]*
         (mydat$f2[mydat$v=="u" & mydat$motID==thismom] - mydat$f2[mydat$v=="a" & mydat$motID==thismom]) +
        mydat$f1[mydat$v=="u" & mydat$motID==thismom]*
         (mydat$f2[mydat$v=="i" & mydat$motID==thismom] - mydat$f2[mydat$v=="a" & mydat$motID==thismom]) +
        mydat$f1[mydat$v=="a" & mydat$motID==thismom]*
         (mydat$f2[mydat$v=="i" & mydat$motID==thismom] - mydat$f2[mydat$v=="u" & mydat$motID==thismom])
      )/2,
          thismom,thisds )
        )
}

colnames(vs)<-c("vss","mom","ds")
write.csv(vs,"output/vs.csv",row.names=F)

read.csv("output/vs.csv")->vs


pivot_wider(vs,names_from="ds",values_from="vss")->vs_wide
write.csv(vs_wide,"output/vs_wide.csv",row.names=F)

vs %>%
    group_by(ds) %>%
    dplyr::summarize(vss = mean(vss, na.rm=TRUE)) 
```

## Additional information on main analyses

Paired contrasts in vowel space size.

```{r ts}
library(effsize)
t.test(vs$vss[vs$ds=="pds"],vs$vss[vs$ds=="ids"],paired=T)
cohen.d(vs$vss[vs$ds=="pds"|vs$ds=="ids"],vs$ds[vs$ds=="pds"|vs$ds=="ids"])
t.test(vs$vss[vs$ds=="ids"],vs$vss[vs$ds=="ads"],paired=T)
cohen.d(vs$vss[vs$ds=="ads"|vs$ds=="ids"],vs$ds[vs$ds=="ads"|vs$ds=="ids"])
t.test(vs$vss[vs$ds=="pds"],vs$vss[vs$ds=="ads"],paired=T)
cohen.d(vs$vss[vs$ds=="pds"|vs$ds=="ads"],vs$ds[vs$ds=="pds"|vs$ds=="ads"])

```

## Duration and pitch analyses


```{r}
mymod=lmer(Vowel.Duration~ds*target + (1 + ds|motID),data=dat)
summary(mymod)

dat$Pitch=as.numeric(as.character(dat$Pitch))
mymod=lmer(Pitch~ds*target + (1+ds|motID),data=dat)
summary(mymod)

```

## F1 and F2 analyses

```{r}
mymod=lmer(f1~ds*target + (1 + ds|motID),data=dat)
summary(mymod)

mymod=lmer(f2~ds*target + (1 + ds |motID),data=dat)
summary(mymod)
```


## Variability analyses

We first get standard deviation of f1 and f2, separating by mom and register.

```{r}
dat %>%
    group_by(motID,ds,target) %>%
    dplyr::summarize(f1.sd = sd(f1, na.rm=TRUE),
                     f2.sd = sd(f2, na.rm=TRUE)) -> sds

```

We then fit separate regressions to each.

```{r}
summary(lmer(f1.sd~ds*target + (1 |motID),data=sds))

```


```{r}
summary(lmer(f2.sd~ds*target + (1|motID),data=sds))

```


## Figure of the vowel space



```{r fig-all-error}

# The palette with grey:
cbPalette <- c( "#56B4E9","#999999","#E69F00")


p <- ggplot(subset(overall_means_nogroup), aes(y = f2, x = f1, col = factor(ds))) +
  geom_polygon(data = subset(overall_means_nogroup) , 
  aes(
     x=f1 ,
    y=f2 ,
    group=ds, 
   linetype=factor(ds)
  ), fill=NA) + #alpha = 0.3 (if we want fill)
  geom_point() + # scale_y_reverse() + scale_x_reverse() +
  annotate("text", x= 350,y=2800,label="i") +
  annotate("text", x= 850,y=1500,label="a") +
  annotate("text", x= 350,y=1350,label="u") +
      geom_errorbar(aes(ymin = overall_means_nogroup$f2-overall_sds_nogroup$f2/sqrt(10),
                         ymax = overall_means_nogroup$f2+overall_sds_nogroup$f2/sqrt(10))) + 
    geom_errorbarh(aes(xmin = overall_means_nogroup$f1-overall_sds_nogroup$f1/sqrt(10),
                      xmax = overall_means_nogroup$f1+overall_sds_nogroup$f1/sqrt(10))) + theme(legend.title = element_blank()) +labs(y= "F2 (Hz)", x = "F1 (Hz)") + theme(panel.background = element_blank()) +scale_colour_manual(values=cbPalette)
 
p 

jpeg("output/fig-vss.jpg",res=300,width=8, height=5, units="cm")
p
dev.off()

pdf("output/fig-vss.pdf",width=5, height=5)
p
dev.off()
```

```{r fig-all2}

overall_means_target$target=factor(overall_means_target$target, levels=c("bead","box","ball","boot"), ordered=T)
overall_means_target[order(overall_means_target$target),]->overall_means_target

sds$target=factor(sds$target, levels=c("bead","box","ball","boot"), ordered=T)
sds[order(sds$target),]->sds

p <- ggplot(subset(overall_means_target), aes(x = f1, y = f2, col = factor(ds)))+
  geom_polygon(data = subset(overall_means_target) ,
  aes(
     x=f1 ,
    y=f2 ,
    group=ds
  ), fill=NA) + #alpha = 0.3 (if we want fill)
  geom_point() 
p 



```
